---
title: "Assignment 1: Effect of Rehabilitation Programs on Mobility in Stroke Patients"
author: "Your Name"
format: 
  html:
    toc: true
    toc-depth: 3
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

# configure global options for R code chunks:
# warning = F to suppress the display of warning messages generated during the execution of code chunks
# message = F to suppress the display of messages produced during the execution of code chunks
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = FALSE)
```

# Introduction

*Provide a brief introduction to the dataset and the objective of the analysis.*

# Exploratory Data Analysis

*Perform an exploratory data analysis to understand the distribution of mobility improvement across treatment groups.*

```{r}
#| label: load dataset
#| echo: true

# Ensure your working directory is set correctly so that R can locate the data file.
# You can set the working directory with setwd("path/to/your/directory") if needed,
# or use the RStudio Session > Set Working Directory > To Source File Location menu.
data_mobility_2 <- read.csv("data_mobility.csv")
```

```{r}
#| label: exploratory-analysis
#| echo: true

# Code for exploratory data analysis goes here
```

# One-Way ANOVA

## Model Estimation

*Fit a one-way ANOVA model.*

```{r}
#| label: anova-model
#| echo: true

# Code for fitting the ANOVA model goes here
```

## ANOVA Table

*Generate the ANOVA table to assess the significance of the treatment effect.*

```{r}
#| label: anova-table
#| echo: true

# Code to generate the ANOVA table
```

# Post-Hoc Analysis

*If significant differences are found, conduct post-hoc comparisons.*

```{r}
#| label: post-hoc
#| echo: true

# Code for post-hoc analysis
```

# Model Diagnostics

*Assess the model fit by generating diagnostic plots.*

```{r}
#| label: diagnostics
#| echo: true

# Code for diagnostics
```

# Conclusion

*Summarize the findings of your analysis and any conclusions about the treatment effects on mobility improvement.*
