---
title: "Beyond MLR Lab 4: Replicated Randomized Block Designs"
format: 
  html:
    toc: true       # Enable the table of contents
    toc-depth: 3    # Set the depth of headers included in the ToC (e.g., H1, H2, H3)
    toc-location: right   # Optional: Can be left, right, or floating    
  pdf:
    toc: false
---

```{r}
#| label: setup
#| include: false

# configure global options for R code chunks:
# warning = F to suppress the display of warning messages generated during the execution of code chunks
# message = F to suppress the display of messages produced during the execution of code chunks
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = FALSE)

# This qmd file requires the following packages to be installed:
#c("dplyr", "ggplot2", "emmeans", "lme4", "lmerTest")
```
In the previous lab, we considered a traditional randomized block design without replication, where each treatment was applied once within each block (laboratory). This design allowed us to control for variability between laboratories and efficiently estimate the **main effects** of the treatments. However, a key limitation of this approach is that it does not allow us to assess whether the treatment effects vary across blocks, a concept known as **interaction**.

The lack of replication in the traditional design means we can only test for main effects of the treatments and laboratories, but not whether the effectiveness of a treatment depends on the specific laboratory conditions. For example, a treatment might work well in one laboratory but not in another due to environmental differences (e.g., humidity or temperature), indicating a **treatment-by-laboratory interaction**.

An interaction occurs when the effect of one factor (e.g., treatment) is not consistent across levels of another factor (e.g., laboratory). In other words, the effect of a treatment might vary depending on the laboratory conditions. Replicating the treatments within each block allows us to estimate these interaction effects and better understand the variability in treatment outcomes across different settings.

In this lab, we introduce replication within blocks, allowing us to estimate the variability within treatment-block combinations and assess whether treatment effects are consistent across blocks.

# Example: Testing Wound Healing Treatments with Replication and Treatment-by-Block Interactions

To allow for replication and potential interaction between treatments and laboratories, we update the simulation code from the previous lab and run it to create an updated dataset: 

```{r}
# Set seed for reproducibility
set.seed(123)

# Define blocks (laboratories) and treatments
blocks <- factor(paste0("Laboratory ", rep(1:5, each = 6)))  # Labs 1 to 5, now with 6 observations per lab
treatments <- factor(rep(c("A", "B", "C"), each = 2, times = 5))  # Replication within blocks

# Simulate block effects and treatment effects
block_effect <- rnorm(5, mean = 0, sd = 2)  # Random effect for each block
treatment_effect <- c(A = 5, B = 7, C = 6)  # Fixed effects for treatments

# Simulate interaction terms (Treatment by Laboratory)
interaction_effect <- matrix(rnorm(15, mean = 0, sd = 1), nrow = 5, ncol = 3)

# Create data frame
data_rep_block <- data.frame(
  Laboratory = blocks,
  Treatment = treatments,
  WoundHealing = NA
)

# Assign responses with interaction terms
for (i in 1:nrow(data_rep_block)) {
  b <- as.numeric(data_rep_block$Laboratory[i])
  t <- data_rep_block$Treatment[i]
  t_index <- match(t, c("A", "B", "C"))
  data_rep_block$WoundHealing[i] <- 
    50 + block_effect[b] + treatment_effect[t] + interaction_effect[b, t_index] + rnorm(1, mean = 0, sd = 1)
}
```

The R code chunk above simulates wound healing observations (measured on a continuous scale, where higher values indicate better healing) for 30 cell culture plates, divided across five laboratory facilities. Each treatment (A, B, and C) is applied to two cell culture plates within each laboratory, introducing replication. The results are stored in the data frame `data_rep_block`, which consists of the following three variables:

- `Laboratory`: A factor variable indicating the laboratory facility.
- `Treatmment`: A factor variable indicating the applied treatment. 
- `WoundHealing`: A numeric variable representing the degree of wound healing on a continuous scale.

## Exploratory Data Analysis

To explore the data and investigate the presence of a potential interaction between treatments and laboratories, we create an  **interaction plot**. An interaction plot is a graphical tool that helps visualize whether the effect of one factor (in this case, treatment) depends on the levels of another factor (in this case, laboratory). Specifically, it shows how the mean response (wound healing) for each treatment varies across different laboratories. If the lines on the interaction plot are parallel, this suggests there is no interaction, meaning the treatment effect is consistent across laboratories. If the lines are not parallel, this indicates a potential interaction, meaning the effect of treatment differs depending on the laboratory.

```{r}
# Calculate mean Wound Healing for each Treatment-Laboratory combination
library(dplyr)
mean_data <- data_rep_block %>%
  group_by(Laboratory, Treatment) %>%
  summarise(Mean_WoundHealing = mean(WoundHealing))

# Display the produced summary table with the group means
mean_data

# Create an interaction plot
library(ggplot2)
ggplot(mean_data, aes(x = Treatment, y = Mean_WoundHealing, group = Laboratory, color = Laboratory)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Interaction Plot",
       y = "Mean Wound Healing Measure",
       x = "Treatment") +
  theme_minimal()
```

::: {.callout-important icon=false title="Question"}
Does the interaction plot indicate a potential interaction between treatment and laboratory?
:::

::: {.callout-important icon=false title="Question"}
Does the plot suggest any differences in the overall effectiveness of the treatments (i.e., main effect of treatment) across all laboratories?
:::

::: {.callout-important icon=false title="Question"}
What does the plot suggest about variability between laboratories? Are some laboratories consistently higher or lower in wound healing across all treatments?
:::

## A Mixed Effects Model for the Randomized Block Design

Similar to the previous lab, laboratory is included as a random effect and treatment as a fixed effect in the model. The key new feature is the **treatment-by-laboratory interaction** term, which is also modeled as a random effect.

To understand why the interaction term must also be modeled as random, we need to consider the nature of random effects in this context. Since laboratory is treated as a random effect, we are assuming that the laboratories in the experiment represent a random sample from a larger population of possible laboratory conditions. This means that we are not just interested in the specific laboratories in the study, but in how the treatments would perform across any set of laboratories with varying conditions.

When we include an interaction term between treatment and laboratory, we are asking whether the effect of each treatment depends on the laboratory environment. If we had modeled laboratory as a fixed effect (i.e., we were only interested in those specific laboratories), the interaction could also be treated as fixed. However, because laboratory is random, the interaction must also be treated as random to reflect the idea that the variability in treatment effects across laboratories applies not just to the specific laboratories in the study, but to any laboratory from the broader population.

In other words, by modeling the interaction as random, we are assuming that the variation in treatment effects is not unique to the five laboratories in the study, but rather represents random fluctuations in treatment effectiveness that could occur in any laboratory setting. This approach allows us to generalize our findings beyond the laboratories in the experiment, making the model more realistic and applicable to a wider range of scenarios.

#### Model Specification

The mixed-effects model for the replicated randomized block design is specified as:

$$
Y_{ijk} = \mu + \tau_i + b_j + (tb)_{ij} + \epsilon_{ijk}
$$

where:

- $Y_{ijk}$: Response (wound healing) for the $k$-tjh replicate of Treatment $i$ in Laboratory $j$.
- $\mu$: Overall mean response
- $\tau_{i}$: Fixed effect of Treatment $i$.
- $b_{j}$:Random effect of Laboratory $j$, assumed to follow a normal distribution with mean zero and variance $\sigma^2_{b}$.
- $(tb)_{ij}$: Random interaction effect between treatment $i$ and Laboratory $j$, assumed to follow a normal distribution with mean zero and variance $\sigma^2_{tb}$.
- $\epsilon_{ijk}$: Random error term, assumed to follow a normal distribution with mean zero and variance $\sigma^2$.

### Model Estimation

Similar as in the previous lab, we fit the mixed effects model using the `lmer()` function from the `lmeTest` package:

```{r}
library(lmerTest)

options(contrasts = c("contr.sum", "contr.poly"))  # Effects coding

model_block <- lmer(WoundHealing ~ Treatment + (1 | Laboratory) + (1 | Treatment:Laboratory), data = data_rep_block)
summary(model_block)
```

#### Syntax overview

Let’s break down the model formula `WoundHealing ~ Treatment + (1 | Laboratory) + (1 | Treatment:Laboratory)`:

- `WoundHealing ~ Treatment`: This specifies that `WoundHealing` is the outcome variable and that `Treatment` is included as a fixed effect to estimate differences between the three treatments.
- `(1 | Laboratory)`: This specifies the random effect for `laboratory`, representing laboratory-specific deviations from the overall mean.
- `(1 | Treatment:Laboratory)`: This specifies the random interaction term between `Treatment` and `Laboratory`.

#### Summary overview

When we run `summary(model_random)`, the following output is displayed:

- **Fixed Effects**
  - The **intercept** represents the overall mean wound healing measure across all treatments and laboratories, which is estimated as 56.47.
  - The coefficient for **Treatment1** (-1.04) indicates that the mean wound healing for Treatment A is 1.04 units lower than the overall mean.
  - The coefficient for **Treatment2** (1.03) shows that the mean wound healing for Treatment B is 1.03 units higher than the overall mean.
  - Treatment C is not directly shown in the output because, with effects coding, the sum of the treatment coefficients must equal zero. Treatment C can be inferred as having a mean very close to the overall mean (around 0.01 units below it).
  
<!-- -->
  
- **Variance components**
  - The random effect for **Laboratory** has a variance of 1.4462, indicating considerable variability in baseline wound healing levels across different laboratories.
  - The random effect for the  **Treatment-by-Laboratory Interaction** has a variance of 0.4747, showing that the treatment effects vary across laboratories. This means the effectiveness of treatments is not entirely consistent across laboratories, though the variability in this interaction is smaller compared to the variability between laboratory baselines and may not be statistically significant (more about that later).
  - The **residual variance** is 0.9089, which represents the unexplained variability within each treatment-laboratory combination.

### Assessing the variance components

In our model, the treatment-by-laboratory interaction is modeled as a random effect. To determine whether this variance component is necessary, we perform a  **likelihood ratio test** to compare the full model (i.e., the previously fitted model with both the random laboratory effect and the random treatment-by-laboratory interaction effect) to a reduced model that includes only the random laboratory effect:

```{r echo=TRUE}
# Fit the full model with the random interaction term
full_model <- lmer(WoundHealing ~ Treatment + (1 | Laboratory) + (1 | Treatment:Laboratory), 
                   data = data_rep_block)

# Fit the reduced model without the random interaction term
reduced_model <- lmer(WoundHealing ~ Treatment + (1 | Laboratory), 
                      data = data_rep_block)

# Perform the likelihood ratio test
anova(full_model, reduced_model, refit=FALSE)
```

When given two or more arguments representing fitted models, the `anova()` function from the `lmerTest` package produces likelihood ratio tests to compare these models. By default, these models are refitted using maximum likelihood (ML) estimation, which is suitable for testing fixed effects but not ideal for testing random effects. Random effects are better assessed using restricted maximum likelihood (REML). 

To keep the REML estimation intact for testing random effects, we use the `refit=FALSE` argument in the `anova()` function. This ensures that the models are compared without being refitted using ML.

The output provides the likelihood ratio test statistic and p-value, indicating whether the random interaction term significantly improves the model fit.

::: {.callout-important icon=false title="Question"}
Based on the results of the likelihood ratio test, is it necessary to include the random interaction term in the model, or is the simpler model (without the interaction term) sufficient?
:::

::: {.callout-note title="Maximum Likelihood (ML) versus Restricted Maximum Likelihood (REML) estimation" collapse="true"}

- **Maximum Likelihood (ML)**: This method estimates all model parameters simultaneously, including both fixed and random effects. It provides unbiased  estimates for fixed effects but can lead to biased estimates of the variance components.

- **Restricted Maximum Likelihood (REML)**: REML focuses on estimating variance components while accounting for the loss of degrees of freedom associated with fixed effects. This method provides more accurate estimates of random effects.

Estimates of the fixed effects parameters are generally comparable between ML and REML. However, variance component estimates will typically be larger with REML given that the method was developed to overcome the downward bias of the maximum likelihood estimates of variance components. For this reason, REML is the default estimation approach when fitting mixed effects models with the `lmer()` function. However, there are also situations where REML cannot be used, particularly during the model-building phase when comparing models with different fixed effects structures. In these cases, the models need to be refitted using ML. 
:::

### Assessing the fixed effects

The fixed effects can be assessed using an approach similar to that used in the previous lab. This includes testing for the overall significance of the treatment effect and, if significant, exploring the differences between specific treatment groups through estimated marginal means and pairwise comparisons.

However, before performing these tests, we need to determine which model to use: the full model with the random interaction term or the reduced model without it. This choice is based on the results of the likelihood ratio test performed earlier. If the likelihood ratio test indicates that the random interaction term is necessary, we will use the full model for testing the fixed effects. Conversely, if the likelihood ratio test suggests that the random interaction term is unnecessary, we will use the simpler model without it. This aligns with the general scientific principle of preferring more parsimonious models that adequately explain the data while minimizing complexity.

### Model diagnostics

The process of performing model diagnostics for the replicated randomized block design is the same as for the traditional randomized block design. Therefore, we refer to the previous lab for detailed instructions on assessing model assumptions and performing diagnostic checks.
