---
title: "Beyond MLR Lab 5: multilevel analysis"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    code-overflow: wrap
    embed-resources: true 
  pdf: 
    toc: false
execute:
  warning: false
  message: false
  eval: false
  echo: true
---

::: {.callout-note icon=false title="Preliminary setup"}
In this lab, we will use the R packages `ggplot2`, `dplyr`, `emmeans`, and `lmerTest`. You can use the script below to automatically install and load them using the `pacman` package.

```{r}
# Check whether pacman is available and install if needed
options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

# Use pacman to install (if needed) and load the required packages
pacman::p_load(dplyr, ggplot2, emmeans, lmerTest)
```
:::

In this lab, we will explore the analysis of multilevel data, focusing on situations where subjects are nested within larger contexts, such as schools, neighborhoods, or clinics. The data include subject-level characteristics (e.g., age, gender), which reflect differences between individuals and help explain variation in their outcomes within a specific context. They also include context-level characteristics (e.g., school funding, neighborhood income), which account for differences between contexts and help explain variation in subject-level outcomes across these contexts.

# Chicago AirBnB Data

The Chicago Airbnb dataset was compiled by Trinh and Ameri as part of a course project at St. Olaf College and is included with the book Beyond Multiple Linear Regression by Roback and Legler (2021). It contains information on 1,561 Airbnb listings in Chicago, including details such as nightly price, overall satisfaction rating, number of reviews, and various other listing characteristics. Additionally, the dataset includes neighborhood-level information, such as ratings for walkability, access to public transit, and bikeability, providing valuable context for the listings' locations.

The dataset is available on Brightspace under the name `airbnb.csv`. It contains the following variables:

* price: the nightly price of the listing (in USD)
* overall_satisfaction: the listing's average rating, on a scale from 1 to 5
* reviews: number of user reviews the listing has
* room_type: the type of listing (eg: Shared room)
* accommodates: number of guests the listing accommodates
* bedrooms: the number of bedrooms the listing has
* minstay: the minimum number of nights to stay in the listing
* neighborhood: the neighborhood in which the listing is located
* district: the broader district in which the listing is located
* WalkScore: the neighborhood's rating for walkability (0 - 100)
* TransitScore: the neighborhood's rating for access to public transit (0 - 100)
* BikeScore: the neighborhood's rating for bikeability (0 - 100)

## Exploratory Data Analysis

Let's load the data and use the `str()` function to inspect the structure of the dataset.

```{r}
# Load the Chicago AirBnB data. Note that the call below assumes the csv file is placed in the 'data' folder directly above the root folder of the project. Update the string as needed if the file is located elsewhere.
chicago_airbnb <- read.csv("data/airbnb.csv")

# Convert all character variables into factors
chicago_airbnb <- chicago_airbnb %>%
  mutate_if(is.character, as.factor)

# Inspect the structure of the dataset
str(chicago_airbnb)
```

This initial inspection makes clear that the Airbnb dataset has a multilevel structure: 1,561 individual listings are nested within 43 neighborhoods, which are further nested within 9 districts. To get a better feeling for this grouping structure, we begin by summarizing the number of listings in each neighborhood:

```{r}
# Summarize number of listings within each neighborhood
listing_summary <- chicago_airbnb %>%
  group_by(neighborhood) %>%
  summarise(num_listings = n())

# Visualize the distribution of listings across neighborhoods
ggplot(listing_summary, aes(x = reorder(neighborhood, num_listings), y = num_listings)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  coord_flip() +
  labs(title = "Number of Listings per Neighborhood",
       x = "Neighborhood",
       y = "Number of Listings")
```

Explanation:

* `group_by(neighborhood)`: groups the data by the neighborhood column so that operations like counting can be applied to each group separately.
* After grouping, `summarise()` calculates the number of rows (listings) in each group (neighborhood) using the `n()` function. The result is stored in a new data fame with two columns:
  - `neighborhood`: the name of the neighborhood;
  - `num_listings`: the number of listings in that neighborhood.

::: {.callout-important icon=false title="Question"}
What does the plot tell us about the distribution of Airbnb listings across neighborhoods?
:::

::: {.callout-important icon=false title="Coding excerise"}
Create a similar plot to visualize the distribution of neighborhoods across districts. Hint: group the dataset by the `district` column and count the number of unique neighborhoods in each district. See `?summarise` for a list of useful functions to use within `summarise()`.
:::

Next, we create a histogram to visualize the distribution of the outcome variable `price`:

```{r}
# Create a histogram of price
ggplot(chicago_airbnb, aes(x = price)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "Histogram of Listing Prices",
       x = "Price",
       y = "Frequency")
```

While normality of the outcome is not strictly required for a mixed-effects model, transforming a right-skewed variable like price can help stabilize variance and linearize relationships. We therefore create a new variable, `log_price`, to store the log-transformed prices:

```{r}
# Log-transform the price variable
chicago_airbnb <- chicago_airbnb %>%
  mutate(log_price = log(price))
```

We also want to get a sense of the variability in the listing prices within and between neighborhoods. For this, we are going to select the 10 neighborhoods with the most listings and create a scatter plot with the log-transformed price on the y-axis and the neighborhood on the x-axis:

```{r}
# Select the top 10 neighborhoods with the most listings
top_neighborhoods <- listing_summary %>%
  slice_max(num_listings, n = 10) %>%
  pull(neighborhood)

# Filter the data to include only the top neighborhoods
top_neighborhood_data <- chicago_airbnb %>%
  filter(neighborhood %in% top_neighborhoods)
```

Explanation:

* `slice_max(num_listings, n = 10)`: selects the top 10 neighborhoods with the most listings based on the `num_listings` column.
* `pull(neighborhood)`: extracts the neighborhood names from the resulting tibble as a vector.
* `filter(neighborhood %in% top_neighborhoods)`: filters the `chicago_airbnb` dataset to include only rows where the neighborhood column matches one of the neighborhoods in the `top_neighborhoods` vector.

```{r}
# Calculate the average log_price for each neighborhood
neighborhood_avg_price <- top_neighborhood_data %>%
  group_by(neighborhood) %>%
  summarise(avg_log_price = mean(log_price, na.rm = TRUE))

# create a scatter plot of log_price by neighborhood
ggplot(top_neighborhood_data, aes(x = neighborhood, y = log_price)) +
  geom_point(alpha = 0.5) +
  labs(title = "Price by Neighborhood",
       x = "Neighborhood",
       y = "Log-transformed price") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_point(data = neighborhood_avg_price, aes(x = neighborhood, y = avg_log_price), color = "red", size = 3) +  # neighborhood-level averages 
  geom_hline(yintercept = mean(neighborhood_avg_price$avg_log_price, na.rm = TRUE), linetype = "dashed", color = "blue") # overall average
```

::: {.callout-important icon=false title="Question"}
What does this plot suggest about the variability in listing prices within and between neighborhoods?
:::

## Random intercept model 

To model the variability in listing prices within and between neighborhoods, we start by fitting a random intercept model to account for the nesting of listings within neighborhoods, ignoring the higher-level nesting of neighborhoods into districts. The model is specified as follows:

```{r}
# Fit the random intercept model
random_intercept_model <- lmer(log_price ~ 1 + (1 | neighborhood), data = chicago_airbnb)
summary(random_intercept_model)
```

::: {.callout-important icon=false title="Question"}
Based on the estimated variance components, what can you conclude about the relative contributions of within-neighborhood and between-neighborhood variability to the total variability in listing prices?
:::

## Extending the random intercept model with subject-level variables

As a second step, we extend the random intercept model with subject-level variables. To facilitate the interpretation of the model coefficients, we start by setting the coding scheme for categorical variables to effects coding. We also center the numerical variables by subtracting the mean value from each observation. This step is important to reduce multicollinearity, improve numerical stability, and make the coefficients more interpretable.

```{r}
# Use effects coding for the categorical variables
options(contrasts = c("contr.sum", "contr.poly"))

# Center the continuous subject-levl variables
chicago_airbnb <- chicago_airbnb %>%
  mutate(overall_satisfaction_c = scale(overall_satisfaction, scale = FALSE),
         accommodates_c = scale(accommodates, scale = FALSE),
         bedrooms_c = scale(bedrooms, scale = FALSE),
         minstay_c = scale(minstay, scale = FALSE))
```

Next, we fit the random intercept model with the subject-level variables included:

```{r}
# Fit the random intercept model with subject-level variables    
random_intercept_model_L1variables <- lmer(log_price ~ overall_satisfaction_c + room_type + accommodates_c + bedrooms_c + minstay_c + (1 | neighborhood), data = chicago_airbnb)
summary(random_intercept_model_L1variables)

# Display the coding scheme for the room_type variable
contrasts(chicago_airbnb$room_type)

# Obtain the ANOVA table for the subject-level variables
anova(random_intercept_model_L1variables)
```

::: {.callout-important icon=false title="Question"}
How does centering the continuous variables affect the interpretation of the intercept?
:::

::: {.callout-important icon=false title="Question"}
How does the inclusion of individual-level variables affect the estimated variance components? More specifically, does the inclusion of individual-level variables affect the within-neighborhood variance, the between-neighborhood variance, or both? Can you explain why?
:::

::: {.callout-important icon=false title="Question"}
Which of the individual-level variables are significantly associated with listing prices? How do you interpret the coefficients for these variables?
:::

## Including context-level variables

As a third step, we extend the previously fitted model with context-level variables. We start by centering the context-level variables:

```{r}
# Center the context-level variables
chicago_airbnb <- chicago_airbnb %>%
  mutate(WalkScore_c = scale(WalkScore, scale = FALSE),
         TransitScore_c = scale(TransitScore, scale = FALSE),
         BikeScore_c = scale(BikeScore, scale = FALSE))
```

Next, we fit the random intercept model with both subject-level and context-level variables included:

```{r}
# Fit the random intercept model with subject-level and context-level variables
random_intercept_model_L1L2variables <- lmer(log_price ~ overall_satisfaction_c + room_type + accommodates_c + bedrooms_c + minstay_c + WalkScore_c + TransitScore_c + BikeScore_c + (1 | neighborhood), data = chicago_airbnb)
summary(random_intercept_model_L1L2variables)

# Obtain the ANOVA table
anova(random_intercept_model_L1L2variables)  
```

::: {.callout-important icon=false title="Question"}
How does the inclusion of context-level variables affect the estimated variance components? More specifically, does the inclusion of context-level variables affect the within-neighborhood variance, the between-neighborhood variance, or both? Can you explain why?
:::

Not all the context-level variables included in the model are significant predictors of listing prices. To obtain a more parsimonious model, we use the `step()` function from the `lmerTest` package to perform a backward elimination of the fixed-effect terms:

```{r}
# Perform stepwise selection to identify the most important predictors
step(random_intercept_model_L1L2variables, reduce.random = FALSE)
```

::: {.callout-important icon=false title="Question"}
Which individual-level and context-level variables are retained in the final model after the stepwise selection procedure? 
:::

## Exploring cross-level interactions

Finally, we explore the possibility of cross-level interactions between individual-level and context-level variables.

The relationship between the overall satisfaction rating of an individual listing and its price may depend on neighborhood characteristics such as their walkability and access to public transit. For instance, higher satisfaction ratings might have a stronger effect on prices in less walkable or transit-accessible neighborhoods, where positive reviews could help compensate for the disadvantages of limited walkability or transit options. In contrast, in neighborhoods with high walkability or access to public transit ratings, these location-based amenities might already drive prices, reducing the added impact of satisfaction ratings.

To test whether the effect of the individual-level variable `overall_satisfaction` on listing prices varies across different values of the context-level variable `WalkScore`, we refit the model with an interaction term between these two variables:

```{r}
# Fit the model with the interaction between overall_satisfaction and WalkScore
random_intercept_model_interaction <- lmer(log_price ~ overall_satisfaction_c * WalkScore_c + room_type + accommodates_c + bedrooms_c + minstay_c + WalkScore_c + TransitScore_c + BikeScore_c + (1 | neighborhood), data = chicago_airbnb)
summary(random_intercept_model_interaction)
```
::: {.callout-important icon=false title="Question"}
Is there evidence of a significant interaction between overall satisfaction ratings and neighborhood walkability in predicting listing prices? If so, how do you interpret the interaction term coefficient?
:::

Similarly, to test whether the effect of the individual-level variable `overall_satisfaction` varies across different values of the context-level variable `TransitScore`, we refit the model with an interaction term between these two variables:

```{r}
# Fit the model with the interaction between overall_satisfaction and WalkScore
random_intercept_model_interaction <- lmer(log_price ~ overall_satisfaction_c * WalkScore_c + room_type + accommodates_c + bedrooms_c + minstay_c + WalkScore_c + TransitScore_c + BikeScore_c + (1 | neighborhood), data = chicago_airbnb)
summary(random_intercept_model_interaction)
```
::: {.callout-important icon=false title="Question"}
Is there evidence of a significant interaction between overall satisfaction ratings and neighborhood walkability in predicting listing prices? If so, how do you interpret the interaction term coefficient?
:::

### Model diagnostics

The process of performing model diagnostics for the random intercept models fitted in this lab is similar to the one described in the previous labs. Therefore, we refer to those labs for more detailed instructions on assessing model assumptions.
