---
title: "Beyond MLR Lab 1: Fixed Effects Model for the One-Way Layout"
format: 
  html:
    toc: true       
    toc-depth: 3    
    toc-location: right   # Optional: Can be left, right, or floating    
  pdf:
    toc: false      
editor_options:
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

# configure global options for R code chunks:
# warning = F to suppress the display of warning messages generated during the execution of code chunks
# message = F to suppress the display of messages produced during the execution of code chunks
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = FALSE)
```

Welcome to lab 1 of the Beyond MLR course. In this lab, we make use of the following R packages: `gglot2`, `dplyr`, and `emmeans`. You can use the script below to check if these packages are installed. Any missing packages can be installed using `install.packages("package_name")`. 

```{r}
#| eval: false
packages <- c("dplyr", "ggplot2", "emmeans")

installed <- sapply(packages, requireNamespace, quietly = TRUE)
if (all(installed)) {
  message("All packages are installed.")
} else {
  missing <- names(installed)[!installed]
  message("The following packages are not installed: ", paste(missing, collapse = ", "))
}
```

# Example: Effect of Three Different Drug Treatments on Blood Pressure Reduction

In this lab, we will examine the effect of three different drug treatments (Control, Drug A, and Drug B) on blood pressure reduction in patients. The data for this example are hypothetical and will be created through simulation:

```{r}
# Generating data for the single-factor experimental design
set.seed(123)
control <- rnorm(40, mean = 2, sd = 10)
drugA <- rnorm(40, mean = 20, sd = 10)
drugB <- rnorm(40, mean = 5, sd = 10)

# Combining data into a data frame
data_oneway <- data.frame(
  Treatment = factor(rep(c("Control", "DrugA", "DrugB"), each = 40),
                     levels = c("Control", "DrugA", "DrugB")),
  BP_change = c(control, drugA, drugB)
)
```

The above R chunk simulates hypothetical results for 120 patients (40 per treatment group) and stores the results in the data frame `data_oneway` that consists of the following two columns:

* `Treatment`: Factor variable indicating the treatment group.
* `BP_change`: Numeric variable representing blood pressure reduction (in mm Hg).

## Exploratory Data Analysis

To understand the distribution of blood pressure reduction across treatment groups, we start by creating a boxplot:

```{r}
# Visualizing the results
library(ggplot2)
ggplot(data_oneway, aes(x = Treatment, y = BP_change, fill = Treatment)) +
  geom_boxplot() +
  labs(title = "Blood Pressure Reduction by Treatment",
       x = "Treatment", y = "Blood Pressure Reduction (mm Hg)") +
  theme_minimal() +
  theme(legend.position = "none")
```

We also calculate some summary statistics:

```{r}
# Summarizing data
library(dplyr)
summary_stats <- data_oneway %>%
  group_by(Treatment) %>%
  summarise(
    Mean_BP_change = mean(BP_change),
    SD_BP_change = sd(BP_change)
  )

summary_stats
```

::: {.callout-important icon=false title="Question"}
How do the mean blood pressure reductions compare among the treatment groups?
:::

## Performing a One-Way ANOVA

### Model Specification

Using effects coding, the one-way Analysis of Variance (ANOVA) model can be specified as:


$$
Y_{ij} = \mu + \tau_i + \epsilon_{ij}
$$

where:

- $Y_{ij}$: The blood pressure reduction for the $j$-th patient in the $i$-th treatment group.
- $\mu$: The overall mean blood pressure reduction.
- $\tau_i$: The effect of the $i$-th treatment (deviation from the overall mean).
- $\epsilon_{ij}$: The error term, assumed to be independently and normally distributed with mean zero and constant variance.

::: {#nte-codings .callout-note title="Coding schemes for categorical variables"}
In regression models, categorical variables are represented using coding schemes that allow their inclusion as explanatory variables. In R, dummy coding is the default coding scheme for nominal variables, where each level is compared to a reference category. For ordinal variables (i.e., ordered factors), R uses orthogonal polynomials as the default coding scheme, which can capture linear or nonlinear trends across ordered levels.

While dummy coding is widely used in multiple linear regression models, effects coding is more common in ANOVA models, as it enables interpretation of group differences relative to the grand mean rather than a single reference category. This interpretative difference makes effects coding particularly useful for examining group-level effects in experimental designs. 

To change the coding scheme globally to effects coding:
```{r}
#| eval: false

# Set contrasts to effects coding globally
options(contrasts = c("contr.sum", "contr.poly"))  # Effects coding
```

To reset the coding scheme to the default dummy coding:
```{r}
#| eval: false

# Reset contrasts to default dummy coding
options(contrasts = c("contr.treatment", "contr.poly"))  # Dummy coding
```
:::

### Model Estimation

We can fit the one-way ANOVA model using the `lm()` function:

```{r}
options(contrasts = c("contr.sum", "contr.poly"))  # Effects coding

# Fitting the model with effects coding
model_oneway <- lm(BP_change ~ Treatment, data = data_oneway)
summary(model_oneway)
```

::: {.callout-important icon=false title="Question"}
How do the estimated coefficients relate to the group means?
:::

### ANOVA Table

To test whether the effect of treatment is statistically significant, we use the `anova()` function to obtain the ANOVA table:

```{r}
# ANOVA table
anova_results <- anova(model_oneway)
anova_results
```

::: {.callout-important icon=false title="Question"}
What does the F-test tell us about the treatment effect?
:::

### Estimated Marginal Means

Estimated marginal means, also known as least-squares means, are model-based means that represent the predicted (or expected) response at each level of a factor, averaged over the levels of other variables in the model.

In the context of a one-way ANOVA, where there is only one factor (e.g., treatment group), the estimated marginal means are equal to the observed group means. In more complex models with multiple factors, estimated marginal means provide predicted group means that are averaged over the levels of these additional factors. For example, in a model with both treatment and age as factors, the estimated marginal means for the treatment groups show the treatment means averaged over age, illustrating what these group means are expected to be at specific values of age (or averaged over a grid of age values).

We can calculate the estimated marginal means using the `emmeans()` function from the `emmeans` package:

```{r}
# Obtain estimated marginal means
library(emmeans)
emms <- emmeans(model_oneway, ~ Treatment)
emms
```

::: {.callout-important icon=false title="Question"}
How do the estimated marginal means compare to the observed group means calculated during the exploratory data analysis?
:::

### Pairwise comparisons

Since we found a significant treatment effect, we will conduct pairwise comparisons of the estimated marginal means to identify which specific treatment groups differ significantly from one another. We perform these pairwise comparisons using the `contrast()` function from the `emmeans` package. To account for the increased risk of Type I error due to multiple comparisons, we apply the Bonferroni correction to adjust the p-values, ensuring that the overall significance level remains controlled.

```{r}
# Performing post-hoc analysis with emmeans
contrast(emms, method="pairwise", adjust="Bonferroni")

```

::: {.callout-important icon=false title="Question"}
Based on the results of the pairwise comparisons, which treatment groups differ significantly?
:::

### Model Diagnostics

To assess the adequacy of the fitted model, we create two diagnostic plots:

- **Residuals vs Fitted Plot**: Used to assess homoscedasticity (constant variance) for the errors. A random scatter of residuals around zero indicates equal variance.
- **Normal Q-Q Plot**: Used to assess normality of the errors. Residuals following the reference line suggest normally distributed errors.



```{r}
# Residuals vs Fitted
plot(model_oneway, which = 1, main = "Residuals vs Fitted")

# Normal Q-Q
plot(model_oneway, which = 2, main = "Normal Q-Q")
```

::: {.callout-important icon=false title="Question"}
Do the diagnostic plots suggest any violations of the model assumptions?
:::

<!-- 
### Reporting
> A one-way ANOVA was conducted to compare the effect of three treatments on blood pressure reduction. There was a statistically significant effect of treatment on blood pressure reduction [F(2, 117) = 43.71, p < 0.001]. Estimated marginal means indicated that Drug A (M = 19.93 mm Hg, 95% CI = [17.11 mm Hg, 22.76 mm Hg]) resulted in significantly greater blood pressure reduction than both the Control group (M = 2.45 mm Hg, 95% CI = [-0.37 mm Hg, 5.28 mm Hg], p < 0.001) and Drug B (M = 5.08 mm Hg, 95% CI = [2.26 mm Hg, 7.90 mm Hg], p < 0.001). There was no significant difference between the Control group and Drug B (p = 0.586). P-values were adjusted for multiple testing using the Bonferroni correction.
--> 

