---
title: "Beyond MLR - Week 1: Experimental Design I"
subtitle: "One-way FE/RE, ANOVA, coding, ICC"
author: "Dr. Douwe Postmus"
format:
  revealjs:
    theme: simple
    slide-number: true
    toc: false
    toc-depth: 2
    slide-level: 2
    incremental: false
    code-overflow: wrap
    scrollable: true
    smaller: true
    width: 1280
    height: 720
    margin: 0.06
    max-scale: 2
    css: slides.css
execute:
  warning: false
  message: false
  eval: true
  echo: false
---

## Introduction

- Welcome slide (to be completed later)

## Experimental research

- Experiment: study in which one or more factors are deliberately varied by assigning specific factor levels to experimental units in order to observe the effect on a response under controlled conditions

- Terminology:
  - Factor: variable the researcher controls or manipulates (e.g., drug, dose)
  - Levels (treatments): different values or categories of a factor (e.g., A, B, C; 10 mg, 20 mg)
  - Experimental unit: smallest unit that can be independently assigned a treatment
  - Response: outcome measured for each unit (e.g., blood pressure change)
  - Effect: difference in outcome caused by a change in the factor

## What do we mean by "effect"?

- An effect describes how the response changes when the level of a factor changes
- Effects can be expressed as differences (e.g., mean or risk difference) or ratios (e.g., risk or odds ratio)
- In an experiment, factors are deliberately varied to study their influence on the response
- The goal of experimental design is to ensure that:

  - the observed effects reflect the true influence of the factors rather than other sources of variation
  - these effects can be estimated precisely and efficiently, by minimizing unexplained variation through proper design choices

## Sources of unwanted variation

- Nuisance variables: influence the response but are not of primary interest (e.g., age, litter)
- They can:

  - increase random variation → more noise, less precision
  - cause systematic bias → distorted estimates of effects
- Types of bias if not controlled

  - Selection bias (confounding): systematic baseline differences between groups
  - Performance bias: different care or handling across treatments
  - Detection bias: outcome assessment influenced by knowledge of treatment

## Principles of experimental design

- Control: keep conditions as similar as possible except for the factors of interest
- Randomization: assign units to factor levels by chance → balances known and unknown nuisance variables
- Replication: include multiple experimental units per treatment — enables estimation of variability and statistical testing
- Blocking: group units by nuisance variables (e.g., litter, cage) and randomize within blocks — isolates treatment effects from nuisance effects

## Case Study — Antihypertensive Treatments

- Research question: Do two experimental drugs (Drug A, Drug B) reduce systolic blood pressure more than Control over 12 weeks?
- Design narrative: 120 patients with hypertension are randomly assigned to one of three groups (n = 40 each): Control, Drug A, or Drug B. The study is conducted under comparable conditions; outcome assessors are blinded to allocation.
- Outcome (response): change in systolic BP (mmHg) from baseline to week 12 (positive values indicate reduction)
- Factor (one-way): Treatment with three levels (Control, Drug A, Drug B)
- Experimental units: individual patients; one outcome per patient

```{r}
#| echo: false
# Prepare example data once for subsequent slides (hidden)
set.seed(123)
control <- rnorm(40, mean = 2,  sd = 10)
drugA   <- rnorm(40, mean = 20, sd = 10)
drugB   <- rnorm(40, mean = 5,  sd = 10)
df_bp <- data.frame(
  Treatment = factor(rep(c("Control","DrugA","DrugB"), each = 40),
                     levels = c("Control","DrugA","DrugB")),
  BP_Reduction = c(control, drugA, drugB)
)
```

## Exploratory Data Analysis

```{r}
#| echo: false
boxplot(BP_Reduction ~ Treatment, data = df_bp,
        col = c("gray85","skyblue","lightgreen"),
        ylab = "BP reduction (mmHg)", main = "BP reduction by treatment")
```

```{r}
#| echo: false
aggregate(BP_Reduction ~ Treatment, data = df_bp,
          FUN = function(x) c(Mean = round(mean(x),2), SD = round(sd(x),2)))
```

## Statistical model specification

To compare average BP reduction across the three treatment groups (a single categorical factor), we use the one-way ANOVA model:

$$
\text{BP_Reduction}_{ij} = \mu + \tau_i + \varepsilon_{ij}
$$

- Model components:
  - $\text{BP_Reduction}_{ij}$: blood pressure reduction for the $j$-th patient in the $i$-th treatment group.
  - $\mu$ (grand mean): the mean of the group-level means.
  - $\tau_i$ (treatment deviations): systematic deviation for treatment $i$; constrained so $\sum_i \tau_i = 0$.
  - $\varepsilon_{ij}$ (error): residual for patient $j$ in group $i$.
- Statistical assumptions:
  - $\varepsilon_{ij}\sim\mathcal{N}(0,\sigma^2)$: the errors $\varepsilon_{ij}$ are normally distributed with common variance $\sigma^2$ across groups (homoscedasticity).
  - The errors are independent

::: {.callout-tip title="Index nesting: groups (i) > members (j)"}
Notational convention: statisticians typically use earlier letters (e.g. $i, k, g$) for factors/levels and later letters (e.g. $j, t, n$) for observational indices — this reflects nesting (i outer, j inner) and makes summation limits natural (sum over $j$ for fixed $i$). For example:

$$Y_{ij} = \mu + \tau_i + \varepsilon_{ij}$$

Here $i$ indexes the group (treatment) and $j$ the individual inside that group. So
$Y_{ij}$ is the $j$-th subject in group $i$, and group averages are
$\bar{Y}_{i\cdot} = \frac{1}{n_i}\sum_{j=1}^{n_i} Y_{ij}$.
:::

## Estimating group effects via linear regression

A one-way ANOVA can be written as a linear regression with a categorical predictor, which lets us apply regression tools to estimate and test effects. Using dummy coding with the control group as reference, the statistically equivalent regression model is:

$$
Y_{ij} = \beta_0 + \beta_1 D_{1,ij} + \beta_2 D_{2,ij} + \varepsilon_{ij}
$$

where $D_1, D_2$ are the dummy indicators for DrugA and DrugB (see table below).

| Treatment | Intercept | D1 (DrugA) | D2 (DrugB) |
|---|---:|:---:|:---:|
| Control | 1 | 0 | 0 |
| DrugA   | 1 | 1 | 0 |
| DrugB   | 1 | 0 | 1 |

- Interpretation:
  - Intercept ($\beta_0$): estimated mean for Control
  - $\beta_1$: estimated difference mean(DrugA) − mean(Control)
  - $\beta_2$: estimated difference mean(DrugB) − mean(Control)

- Mapping to the ANOVA model parameters
  - Intercept and betas → group means:
    - mean(Control) = $\beta_0$
    - mean(DrugA)   = $\beta_0 + \beta_1$
    - mean(DrugB)   = $\beta_0 + \beta_2$

  - Grand mean (ANOVA intercept):
    $\mu = \frac{\text{mean(Control)} + \text{mean(DrugA)} + \text{mean(DrugB)}}{3} = \beta_0 + \frac{\beta_1 + \beta_2}{3}$

  - Treatment deviations (ANOVA $\tau_i$) written in terms of the regression coefficients:
    - $\tau_{\text{Control}} = \text{mean(Control)} - \mu = -\frac{\beta_1 + \beta_2}{3}$
    - $\tau_{\text{DrugA}}   = \text{mean(DrugA)} - \mu = \beta_1 - \frac{\beta_1 + \beta_2}{3}$
    - $\tau_{\text{DrugB}}   = \text{mean(DrugB)} - \mu = \beta_2 - \frac{\beta_1 + \beta_2}{3}$

  - In short: the dummy-coded regression coefficients give easy expressions for the group means, and the ANOVA deviations (the $\tau_i$) are simple transforms of the $\beta$s shown above.

## Estimated regression coefficients - dummy coding

```{r}
#| echo: false
# R (refresher): reference coding with Control as baseline
contrasts(df_bp$Treatment) <- contr.treatment(levels(df_bp$Treatment))
fit <- lm(BP_Reduction ~ Treatment, data = df_bp)
summary(fit)
```

## Effects coding and mapping to ANOVA parameters

Another common choice is effects (sum-to-zero) coding. With this coding the regression coefficients map directly to the ANOVA parameters:

$$
Y_{ij} = \beta_0 + \beta_1 E_{1,ij} + \beta_2 E_{2,ij} + \varepsilon_{ij}
$$

where $E_1, E_2$ are the effect indicators for Control and DrugA (see table below).

| Treatment | Intercept | E1 | E2 |
|---|---:|:---:|:---:|
| Control | 1 | 1 | 0 |
| DrugA   | 1 | 0 | 1 |
| DrugB   | 1 | -1 | -1 |

- Interpretation:
  - Intercept ($\beta_0$): grand mean ($\mu$)
  - $\beta_1$ = $\tau_{\text{Control}}$ (deviation of Control from grand mean)
  - $\beta_2$ = $\tau_{\text{DrugA}}$ (deviation of DrugA from grand mean)
  - The remaining group's deviation follows from the sum-to-zero constraint: $\tau_{\text{DrugB}} = -\beta_1 - \beta_2$

## Estimated regression coefficients - effects coding

```{r}
#| echo: false
## R: effects (sum-to-zero) coding
contrasts(df_bp$Treatment) <- contr.sum(levels(df_bp$Treatment))
fit2 <- lm(BP_Reduction ~ Treatment, data = df_bp)
summary(fit2)
```

## Global hypothesis testing

- Purpose: test whether any treatment systematically changes the expected response — i.e., whether the factor explains variability beyond random error.

- Hypotheses:
  - Null: $H_0:\ \tau_1=\tau_2=\dots=\tau_k=0$ (all treatment deviations are zero)
  - Alternative: $H_1:$ at least one $\tau_i\ne0$ (some treatment differs)

 - We test this using the F-test shown in the ANOVA table (below).

```{r}
#| echo: false
anova(fit)
```

- If the F is large (p small): reject $H_0$, then follow up with contrasts or pairwise tests to locate differences.
