---
title: "Assignment Week 2: Effectiveness of Exercise Routines on Functional Mobility"
format: html
---

```{r}
#| label: setup
#| include: false

# configure global options for R code chunks:
# warning = F to suppress the display of warning messages generated during the execution of code chunks
# message = F to suppress the display of messages produced during the execution of code chunks
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = FALSE)
```

## Introduction

In this assignment, you will analyze a dataset that evaluates the effectiveness of different exercise routines on improving functional mobility in patients recovering from knee replacement surgery. The study uses a replicated randomized block design to account for individual variability among patients.

Each patient participated in all three exercise routines, and each routine was performed three times. Functional mobility was assessed after each session, with scores ranging from 40 to 80 (higher scores indicate better mobility). The dataset contains 54 observations (6 patients x 3 routines x 3 replications) and is provided in the downloads section below.

## Dataset Description

The dataset consists of the following variables:

- `Patient`: A numeric variable representing individual patients (blocks).
- `ExerciseRoutine`: A factor variable representing the type of exercise routine:
  - **A**: Low-impact walking exercises.
  - **B**: Resistance training exercises.
  - **C**: Balance and flexibility exercises.
- `FunctionalMobility`: A numeric variable (40-80) representing the mobility score measured after each session.

## Objective

Your goal is to determine:

1. Whether there are significant differences in functional mobility scores between the three exercise routines.
2. How much of the variability in mobility scores can be attributed to:
   - Differences between patients.
   - Differences between exercise routines.
   - Residual variability (within patient-treatment combinations).

To complete this analysis, follow the detailed steps outlined below.

## Steps to Complete the Analysis

### 1. Load the dataset

When loading the dataset in R, ensure that the `ExerciseRoutine` variable is treated as a factor. One way to achieve this is by specifying the column type explicitly using the `colClasses` argument in the `read.csv()` function. For example:

```{r}
# Load the dataset and specify ExerciseRoutine as a factor
dataset <- read.csv("functional_mobility.csv", colClasses = c("ExerciseRoutine" = "factor"))
```

### 2. Perform an Exploratory Data Analysis (EDA)
- Visualize the data using an interaction plot.
- Calculate descriptive statistics for each exercise routine and patient.

### 3. Fit a Mixed-Effects Model
- Fit a mixed-effects model with:
  - **Fixed Effects**: `ExerciseRoutine` (to estimate differences between exercise routines).
  - **Random Effects**: 
    - `Patient` (to account for variability between individuals).
    - `Patient × ExerciseRoutine` (to account for potential variability in how patients respond to different routines).
- Test whether the `Patient × ExerciseRoutine` random interaction term is necessary by comparing the full model (with the interaction term) to a reduced model (without the interaction term).
- Based on the results of this test:
  - If the interaction term is significant, proceed with the full model for further analysis.
  - If the interaction term is not significant, proceed with the reduced model.
- Summarize the chosen model:
  - Report the fixed effects and variance components.

### 4. Test for Overall Differences
- Use the selected model to test for overall differences in functional mobility scores across exercise routines.

### 5. Perform Pairwise Comparisons
- If significant overall differences are found, calculate estimated marginal means and perform pairwise comparisons to identify which routines differ.

### 6. Evaluate Model Assumptions
- Assess the model's assumptions (e.g., normality of residuals, homoscedasticity) using diagnostic plots.

## Requirements for the Report

Submit both the **Quarto source file (`.qmd`)** and the **rendered HTML file**. The quarto source file should include all R code and annotations, and the html file should be rendered with code visibility enabled (`echo = TRUE`). 

## Downloads 

[Download dataset](downloads/functional_mobility.csv)


