---
title: "Assignment 4: longitudinal data analysis"
subtitle: "Model answer"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    code-overflow: wrap
    embed-resources: true
execute:
  warning: false
  message: false
  eval: true
  echo: true
---

## Step 1: Load and prepare the data

```{r}
# Load the required libraries
library(dplyr)
library(ggplot2)
library(lmerTest)

# Load the data
asian_data <- read.csv("Assignments/Week 4/downloads/asian_data.csv")

# Convert all character variables into factors
asian_data <- asian_data |>
  mutate(across(where(is.character), as.factor))

# Use effects coding for categorical variables
options(contrasts = c("contr.sum", "contr.poly"))

# Inspect the structure of the dataset
str(asian_data)
```

```{r}
# Center age at 12 months (approximately the middle of the age range)
asian_data <- asian_data |>
  mutate(
    age_c = Age - 12,
    age_c2 = age_c^2,
    age_c3 = age_c^3
  )

# Check the age range
summary(asian_data$Age)
```

## Step 2: Exploratory data analysis

### Spaghetti plot: individual and mean trajectories

```{r}
# Compute the overall mean weight by age (binned for visualization)
mean_trajectory <- asian_data |>
  mutate(age_bin = round(Age)) |>
  group_by(age_bin) |>
  summarise(mean_weight = mean(Weight, na.rm = TRUE), .groups = "drop")

# Create spaghetti plot
ggplot(asian_data, aes(x = Age, y = Weight)) +
  geom_line(aes(group = ChildID), alpha = 0.2, color = "gray60") +
  geom_smooth(method = "loess", se = FALSE, color = "steelblue", linewidth = 1.5) +
  labs(
    title = "Individual and Mean Growth Trajectories",
    subtitle = "Gray lines = individual trajectories; Blue line = smoothed mean",
    x = "Age (months)",
    y = "Weight (grams)"
  ) +
  theme_minimal()
```

**Observations:**

- Clear upward trend: children gain weight as they age
- Growth appears to decelerate at older ages (suggesting a quadratic or higher-order term may be needed)
- Substantial variation in both baseline weight and growth rates between children

### Mean trajectories by gender

```{r}
# Compute mean trajectories by gender
mean_by_gender <- asian_data |>
  mutate(age_bin = round(Age)) |>
  group_by(age_bin, Gender) |>
  summarise(mean_weight = mean(Weight, na.rm = TRUE), .groups = "drop")

# Plot mean trajectories by gender
ggplot(asian_data, aes(x = Age, y = Weight, color = Gender)) +
  geom_line(aes(group = ChildID), alpha = 0.1) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.5) +
  labs(
    title = "Growth Trajectories by Gender",
    subtitle = "Faint lines = individuals; Bold lines = smoothed group means",
    x = "Age (months)",
    y = "Weight (grams)",
    color = "Gender"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Observations:**

- Boys appear to be heavier than girls on average throughout the age range
- The trajectories appear roughly parallel, suggesting a main effect of gender but possibly no interaction with time

## Step 3: Model the time structure

### Step 3.1: Fit a flexible model for time

We start with a model that includes linear, quadratic, and cubic terms for time, along with random intercept and random slope.

```{r}
# Fit the full time model with cubic polynomial and random slope
time_model_full <- lmer(Weight ~ age_c + age_c2 + age_c3 + (age_c | ChildID),
                        data = asian_data,
                        control = lmerControl(optimizer = "bobyqa"))
summary(time_model_full)
```

### Step 3.2: Test the random slope

```{r}
# Fit model with random intercept only
time_model_ri <- lmer(Weight ~ age_c + age_c2 + age_c3 + (1 | ChildID),
                      data = asian_data,
                      control = lmerControl(optimizer = "bobyqa"))

# Compare models using LRT (refit = FALSE for random effects)
anova(time_model_ri, time_model_full, refit = FALSE)
```

**Conclusion:** The random slope is highly significant (p < 0.001), indicating that children differ in their growth rates. We retain the random slope structure.

### Step 3.3: Calculate the ICC

```{r}
# Extract variance components from the random intercept model
vc <- as.data.frame(VarCorr(time_model_ri))
var_intercept <- vc$vcov[1]
var_residual <- vc$vcov[2]

# Calculate ICC
icc <- var_intercept / (var_intercept + var_residual)
cat("Variance between children:", round(var_intercept, 0), "\n")
cat("Variance within children:", round(var_residual, 0), "\n")
cat("ICC:", round(icc, 3), "\n")
```

**Interpretation:** The ICC indicates that approximately `r round(icc * 100)`% of the variance in weight (after accounting for age trends) is due to stable differences between children.

### Step 3.4: Test the fixed effects for time

We test the polynomial terms starting from the highest order, keeping the random slope structure throughout.

```{r}
# Test cubic term
time_model_quad <- lmer(Weight ~ age_c + age_c2 + (age_c | ChildID),
                        data = asian_data,
                        control = lmerControl(optimizer = "bobyqa"))
anova(time_model_quad, time_model_full, refit = TRUE)
```

**Conclusion:** The cubic term is highly significant (p < 0.001). Following the marginality principle, we retain all lower-order terms (linear and quadratic) without further testing. The final time model includes linear, quadratic, and cubic terms.

### Step 3.5: Residual diagnostics

```{r}
# Create diagnostic plot for cubic model
diag_data <- data.frame(
  fitted = fitted(time_model_full),
  residuals = resid(time_model_full)
)

ggplot(diag_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
  labs(
    title = "Residuals vs Fitted Values (Cubic Model)",
    x = "Fitted values",
    y = "Conditional Residuals"
  ) +
  theme_minimal()
```

**Conclusion:** The loess curve lies close to zero across the range of fitted values, indicating that the cubic polynomial adequately captures the growth trajectory.

### Step 3.6: Comparing alternative time models

Not all students may include higher-order polynomial terms. To illustrate the consequences of under-specifying the time structure, we fit models with (i) linear + quadratic terms only and (ii) linear term only, and examine their residual diagnostics.

#### Quadratic model

```{r}
# Diagnostic plot for quadratic model
diag_data_quad <- data.frame(
  fitted = fitted(time_model_quad),
  residuals = resid(time_model_quad)
)

ggplot(diag_data_quad, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
  labs(
    title = "Residuals vs Fitted Values (Quadratic Model)",
    x = "Fitted values",
    y = "Conditional Residuals"
  ) +
  theme_minimal()
```

#### Linear model

```{r}
# Fit linear-only model
time_model_linear <- lmer(Weight ~ age_c + (age_c | ChildID),
                          data = asian_data,
                          control = lmerControl(optimizer = "bobyqa"))

# Diagnostic plot for linear model
diag_data_linear <- data.frame(
  fitted = fitted(time_model_linear),
  residuals = resid(time_model_linear)
)

ggplot(diag_data_linear, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "loess", se = TRUE, color = "blue", fill = "lightblue") +
  labs(
    title = "Residuals vs Fitted Values (Linear Model)",
    x = "Fitted values",
    y = "Conditional Residuals"
  ) +
  theme_minimal()
```

**Observations:**

- **Quadratic model:** The loess curve shows a slight S-shaped pattern, deviating from zero at the extremes. This suggests that the quadratic model does not fully capture the growth trajectory, though the deviation is modest.
- **Linear model:** The loess curve shows a clear curved pattern, substantially deviating from zero. The model systematically over-predicts at low and high fitted values and under-predicts in the middle range. This is a classic sign that the linear time trend is inadequate.

These diagnostic plots demonstrate the importance of checking residuals when modeling longitudinal data. A model that appears to fit well based on significance tests may still fail to capture important features of the growth trajectory.

## Step 4: Add the gender covariate

### Step 4.1: Fit full model with gender

We add gender and its interactions with the time terms.

```{r}
# Fit model with gender main effect and interactions with time
full_model <- lmer(
  Weight ~ (age_c + age_c2 + age_c3) * Gender + (age_c | ChildID),
  data = asian_data,
  control = lmerControl(optimizer = "bobyqa")
)
summary(full_model)
```

### Step 4.2: Simplify the model

We use the `step()` function to perform backward elimination of non-significant fixed effects.

```{r}
# Use step() to simplify fixed effects
step_result <- step(full_model, reduce.fixed = TRUE, reduce.random = FALSE)
print(step_result)

# Extract the final model
final_model <- get_model(step_result)
```

### Step 4.3: Final model summary

```{r}
summary(final_model)
```

## Step 5: Interpretation

### Fixed effects interpretation

```{r}
# Display fixed effects with confidence intervals
confint(final_model, method = "Wald", parm = "beta_")
```

Based on the final model:

1. **Time trends:** The significant linear, quadratic, and cubic terms indicate that children's weight follows a curved growth trajectory. The positive linear term shows overall weight gain, while the negative quadratic term indicates deceleration of growth at older ages.

2. **Gender effect:** The significant main effect of gender indicates that boys and girls differ in their weight. With effects coding, the coefficient represents the deviation from the grand mean.

3. **Gender × time interaction:** If retained in the final model, this would indicate that boys and girls have different growth rates. If eliminated by `step()`, it suggests the trajectories are parallel (same growth pattern, different baseline).
