---
title: "Assignment 4: longitudinal data analysis"
format: 
  html:
    toc: true       # Enable the table of contents
    toc-depth: 3    # Set the depth of headers included in the ToC (e.g., H1, H2, H3)
    toc-location: right   # Optional: Can be left, right, or floating    
    code-overflow: wrap 
---

```{r}
#| label: setup
#| include: false

# configure global options for R code chunks:
# warning = F to suppress the display of warning messages generated during the execution of code chunks
# message = F to suppress the display of messages produced during the execution of code chunks
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = TRUE)

# This qmd file requires the following packages to be installed:
#c("dplyr", "ggplot2", "emmeans", "lme4", "lmerTest")
```

## Exploratory Data Analysis

Let's load the data and use the `str()` function to inspect the structure of the dataset.

```{r}
# Load the required libraries
library(dplyr)
library(ggplot2)
library(lmerTest)

asian_data <- read.csv("downloads/asian_data.csv")

# Convert all character variables into factors
asian_data <- asian_data %>%
  mutate(across(where(is.character), as.factor))

# Inspect the structure of the dataset
str(asian_data)
```

<!-- > str(asian_data) -->
<!-- 'data.frame':	1572 obs. of  5 variables: -->
<!--  $ ChildID    : int  2 2 3 3 3 3 22 22 22 22 ... -->
<!--  $ Age        : num  12.39 27.1 1.77 7.95 12.39 ... -->
<!--  $ Weight     : int  10262 14500 4053 8136 9752 13180 4510 8360 8480 13150 ... -->
<!--  $ BirthWeight: int  3360 3360 3310 3310 3310 3310 2710 2710 2710 2710 ... -->
<!--  $ Gender     : Factor w/ 2 levels "Boy","Girl": 2 2 2 2 2 2 2 2 2 2 ... -->

## Exploratory data analysis

### Individual trajectories of alcohol use

We will start by examining the individual trajectories of Weight over time. One way to achieve this is a facet plot, which displays each individual's trajectory in a separate subplot. In this case, however, the dataset consists of 568 individuals, making it impractical to display all trajectories. We therefore take a random sample of 25 individuals for visualization.

```{r}
# Sample 25 individuals for visualization
set.seed(123)
sample_ids <- sample(unique(asian_data$ChildID), 25)

# Filter the data for the sampled individuals
sample_data <- asian_data %>%
  filter(ChildID %in% sample_ids)

# Create a facet plot of individual trajectories
ggplot(sample_data, aes(x = Age, y = Weight, group = ChildID)) +
  geom_line() +
  geom_point(size = 2, alpha = 0.7) + 
  facet_wrap(~ChildID, scales = "free_y") +
  labs(title = "Individual Trajectories of Weight Use Over Time",
       x = "Age (months)",
       y = "Weight (grams)")
```

# Create a new variable Age_Centered and Age_Centered_Squared
```{r}
asian_data <- asian_data %>%
  mutate(Age_Centered = Age - mean(Age),
         Age_Centered_Squared = Age_Centered^2)
```

# Fit a random slope model with Age_centered and Age_Centered_Squared
```{r}
# Fit a random slope model with Age_centered and Age_Centered_Squared
model <- lmer(Weight ~ Age_Centered + Age_Centered_Squared + (Age_Centered | ChildID), data = asian_data)
summary(model)
```

# Inlcude Gender, BirthWeight and interactions with Age_centered and Age_Centered_Squared in the model
```{r}
# Fit a random slope model with gender, BirthWeight, and interactions with Age_centered and Age_Centered_Squared
model <- lmer(Weight ~ (Age_Centered + Age_Centered_Squared)*(Gender) + (Age_Centered | ChildID), data = asian_data, control = lmerControl(optimizer = "bobyqa"))
summary(model)

model2 <- lmer(Weight ~ (Age_Centered + Age_Centered_Squared)*(Gender) + (Age_Centered | ChildID), data = asian_data, control = lmerControl(optimizer = "bobyqa"))
summary(model2)

plot(model2)
qqnorm(ranef(model2)$ChildID[, "Age_Centered"])
qqline(ranef(model2)$ChildID[, "Age_Centered"])

```

              
