---
title: "Final Assignment: Cardiac Recovery Analysis"
author: "Your Name (Student ID: XXXXXX)"
format:
  html:
    embed-resources: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| label: packages

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

# Use pacman to install (if needed) and load the required packages
pacman::p_load(dplyr, ggplot2, emmeans, lmerTest)
```

```{r}
#| label: setup

# IMPORTANT: Replace 123456 with YOUR student/staff number!
student_id <- 123456

# Load the master dataset
master_data <- read.csv("cardiac_recovery_master.csv")

# Convert all character variables into factors
master_data <- master_data |>
  mutate(across(where(is.character), as.factor))

# Determine your subpopulation based on last two digits
last_two <- student_id %% 100

if (last_two <= 24) {
  my_subpop <- master_data[master_data$sex == "Female", ]
  subpop_name <- "Female patients"
} else if (last_two <= 49) {
  my_subpop <- master_data[master_data$sex == "Male", ]
  subpop_name <- "Male patients"
} else if (last_two <= 74) {
  my_subpop <- master_data[master_data$diabetes == "Yes", ]
  subpop_name <- "Patients with diabetes"
} else {
  my_subpop <- master_data[master_data$diabetes == "No", ]
  subpop_name <- "Patients without diabetes"
}

# Generate your personal subsample within the subpopulation
set.seed(student_id)
sampled_patients <- sample(unique(my_subpop$patient_id),
                           size = round(0.7 * length(unique(my_subpop$patient_id))))

# Create your personal dataset
my_data <- my_subpop[my_subpop$patient_id %in% sampled_patients, ]

# Display your dataset summary
cat("Student ID:", student_id, "\n")
cat("Subpopulation:", subpop_name, "\n")
cat("Number of patients:", length(unique(my_data$patient_id)), "\n")
cat("Number of observations:", nrow(my_data), "\n")
```

## 1. Introduction

*Describe the clinical context of cardiac surgery recovery. Introduce your assigned subpopulation and explain why studying this group is clinically relevant. State your two research questions clearly.*

## 2. Exploratory Data Analysis

*Explore both outcome variables in your subpopulation. Include:*

- *Descriptive statistics and distributions*
- *Spaghetti plots showing individual trajectories*
- *Mean trajectories by rehabilitation program*
- *Brief interpretation of observed patterns*

## 3. Unconditional Growth Models

*Fit unconditional growth models (with time as the only predictor) for both outcomes.*

### Quality of Life

*Report the fixed effects (intercept and slope), random effects (variance components), and ICC. Check model assumptions using residual plots.*

### Walk Distance

*Report the fixed effects (intercept and slope), random effects (variance components), and ICC. Check model assumptions using residual plots.*

## 4. Conditional Growth Models

*Add rehabilitation program and its interaction with time to test your research questions.*

### Quality of Life

*Present your conditional model, interpret the fixed effects, and determine whether the rehabilitation program significantly affects the recovery trajectory.*

### Walk Distance

*Present your conditional model, interpret the fixed effects, and determine whether the rehabilitation program significantly affects the recovery trajectory.*

## 5. Discussion and Conclusions

*Summarize your key findings for both outcomes. Provide a clinical interpretation of the results—what do they mean for patients in your subpopulation? Discuss limitations of your analysis.*
