---
title: "Assignment 1: Effect of Rehabilitation Programs on Mobility in Stroke Patients"
author: "Your Name"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
---


```{r}
#| label: preliminary-setup 
#| include: false

# Use pacman to load the required R packages (feel free to add packages where needed)
options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(knitr, rmarkdown, dplyr, ggplot2, emmeans)
```

```{r}
#| label: load-data 
#| include: false

# Ensure your working directory is set to the location where you stored the data file.
# You can set the working directory with setwd("path/to/your/directory") if needed,
# or use the RStudio Session > Set Working Directory > To Source File Location menu.
data_mobility <- read.csv("data_mobility.csv", colClasses = c(Treatment = "factor"))
```

# Introduction

*Provide a brief introduction to the dataset and the objective of the analysis.*

# Exploratory Data Analysis

*Perform an exploratory data analysis to understand the distribution of mobility improvement across treatment groups.*

```{r}
#| label: exploratory-analysis

# Code for exploratory data analysis goes here
```

# One-Way ANOVA

## Model Estimation

*Fit a one-way ANOVA model.*

```{r}
#| label: anova-model

# Code for fitting the ANOVA model goes here
```

## ANOVA Table

*Generate the ANOVA table to assess the significance of the treatment effect.*

```{r}
#| label: anova-table

# Code to generate the ANOVA table
```

# Post-Hoc Analysis

*If significant differences are found, conduct post-hoc comparisons.*

```{r}
#| label: post-hoc

# Code for post-hoc analysis (if needed)
```

# Model Diagnostics

*Assess the model fit by generating diagnostic plots.*

```{r}
#| label: diagnostics

# Code for diagnostics
```

# Conclusion

*Summarize the findings of your analysis and any conclusions about the treatment effects on mobility improvement.*
