---
title: "Assignment 3: Multilevel Modeling"
format: 
  html:
    toc: true       # Enable the table of contents
    toc-depth: 3    # Set the depth of headers included in the ToC (e.g., H1, H2, H3)
    toc-location: right   # Optional: Can be left, right, or floating    
    code-overflow: wrap 
---

```{r}
#| label: setup
#| include: false

# configure global options for R code chunks:
# warning = F to suppress the display of warning messages generated during the execution of code chunks
# message = F to suppress the display of messages produced during the execution of code chunks
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = TRUE)

# This qmd file requires the following packages to be installed:
#c("dplyr", "ggplot2", "emmeans", "lme4", "lmerTest")
```

## Exploratory Data Analysis

Let's load the data and use the `str()` function to inspect the structure of the dataset.

```{r}
# Load the required libraries
library(dplyr)
library(ggplot2)
library(lmerTest)

ilea_data <- read.csv("downloads/ilea_data.csv")

# Convert all character variables into factors
ilea_data <- ilea_data %>%
  mutate_if(is.character, as.factor)

# Inspect the structure of the dataset
str(ilea_data)
```

To get a better feeling for this grouping structure, we begin by summarizing the number of students within each school:
```{r}
# Summarize number of students within each school
school_summary <- ilea_data %>%
  group_by(School) %>%
  summarise(num_students = n())

# Visualize the distribution of students across schools
ggplot(school_summary, aes(x = reorder(School, num_students), y = num_students)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  coord_flip() +
  labs(title = "Number of Students per School",
       x = "School",
       y = "Number of Students")
```

Next, we create a histogram to visualize the distribution of the outcome variable `ExamScore`:

We also want to get a sense of the variability in the exam scores within and between schools. For this, we are going to select the 10 neighborhoods with the most listings and create a scatter plot with the exam core price on the y-axis and the school on the x-axis:

```{r}
# Select the top 10 schools with the most students
top_schools <- school_summary %>%
  slice_max(num_students, n = 10) %>%
  pull(School)

# Filter the data to include only the top schools
top_school_data <- ilea_data %>%
  filter(School %in% top_schools)
```

```{r}
# Create a scatterplot of the exam scores by school for the 10 schools with the most students
ggplot(top_school_data, aes(x = as.factor(School), y = ExamScore)) +
  geom_point(alpha = 0.5) +
  labs(title = "Exam Scores by School",
       x = "School",
       y = "Exam Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = mean(ilea_data$ExamScore), linetype = "dashed", color = "blue") 
```

## Random intercept model 

```{r}
# Fit the random intercept model
random_intercept_model <- lmer(ExamScore ~ 1 + (1 | School), data = ilea_data)
summary(random_intercept_model)
```

## Extending the random intercept model with subject-level variables

```{r}
# Use effects coding for the categorical variables
options(contrasts = c("contr.sum", "contr.poly"))
```

Next, we fit the random intercept model with the subject-level variables included:

```{r}
# Fit the random intercept model with subject-level variables    
random_intercept_model_L1variables <- lmer(ExamScore ~ Gender + VRBand + (1 | School), data = ilea_data)
summary(random_intercept_model_L1variables)

contrasts(ilea_data$Gender)
contrasts(ilea_data$VRBand)

# Obtain the ANOVA table for the subject-level variables
anova(random_intercept_model_L1variables)
```

## Including context-level variables

As a third step, we extend the previously fitted model with context-level variables. We start by centering the context-level variables:

```{r}
# Center the context-level variables
ilea_data <- ilea_data %>%
  mutate(PercentFSM_c = scale(PercentFSM, scale = FALSE))
```

Next, we fit the random intercept model with both subject-level and context-level variables included:

```{r}
# Fit the random intercept model with subject-level and context-level variables
random_intercept_model_L1L2variables <- lmer(ExamScore ~PercentFSM_c + (1 | School), data = ilea_data)
summary(random_intercept_model_L1L2variables)

# Obtain the ANOVA table
anova(random_intercept_model_L1L2variables)  
```

## Exploring cross-level interactions

Finally, we explore the possibility of cross-level interactions between individual-level and context-level variables.

```{r}
# Fit the model with the interaction between PercentFSM_c and Gender
random_intercept_model_interaction1 <- lmer(ExamScore ~ PercentFSM_c*Gender + (1 | School), data = ilea_data)
summary(random_intercept_model_interaction1)

# Fit the model with the interaction between PercentFSM_c and VRBand
random_intercept_model_interaction2 <- lmer(ExamScore ~ PercentFSM_c*VRBand + (1 | School), data = ilea_data)
summary(random_intercept_model_interaction2)
```
